---
title: "Spatial Data Analysis "
date: "2023-01-30"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

When one wants to explore and visualize global data, the "World" dataset available in the R package tmap can prove to be a valuable resource. This dataset contains information on the political boundaries and attributes of countries and regions around the world. With the help of this dataset, researchers and analysts can generate maps and visualizations that reveal patterns and relationships within global data. The World dataset has been widely used in various fields, including economics, political science, and environmental studies. Its broad scope and accessibility make it a useful tool for anyone interested in examining global trends and patterns.

```{r}
# load the required libraries
library(tmap)
library(sf)
library(tidyverse)
```


## Part 1.1

Analyzing the type of spatial object that the ‘World’ data set is? What is the CRS?
```{r}
# check the type of world dataset
data('World')
class(World)
```

The World dataset is a simple feature (sf) spatial dataframe.

```{r}
# view the dataframe
head(World)
```


```{r}
# check the CRS
st_crs(World)
```
The CRS of the World dataset is World Geodetic System 1984 EPSG:4326 

## Part 1.2

Using the projection argument of `tm_shape()`, create maps of the World data set with two other projections.

```{r}
# create map
tm_shape(World) + 
    tm_polygons() + 
tm_layout("Long lat coordinates (WGS84)", inner.margins=c(0,0,.1,0), title.size=.8)
```

```{r}
# create map with Eckhart IV projection
tm_shape(World, projection="+proj=eck4") + 
    tm_polygons() + 
tm_layout("Eckhart IV projection", inner.margins=c(0,0,.1,0), title.size=.8)
```

```{r}
# create map with ano projection
tm_shape(World, projection="+proj=robin") + 
    tm_polygons() + 
tm_layout("Winkel-Tripel projection", inner.margins=c(0,0,.1,0), title.size=.8)
```


Stating what aspect are preserved with the projections used. Commenting on the visual differences and biases across the three maps.

WGS84 projection: This is an equal-area projection, and it preserves area accurately but distorts shapes and angles.

Eckert IV projection: This is an equal-area projection that displays continents with an elliptical shape and a low amount of distortion. It preserves area accurately but distorts shape and direction.

Winkel-Tripel projection: This projection is an equal-area projection and it has reduced distortion of both shape and area. It has a low amount of distortion and generally shows accurate directions, but is not an exact representation of size or shape.


## Part 1.3

The "World" data set contains a column labeled “HPI” that stands for Happy Planet Index. Information about this metric can be found here: http://happyplanetindex.org The code makes a four panel map that plots: HPI, life expectancy, footprint and inequality.

```{r}
# plot 4 maps together
plot(World[,c("HPI","life_exp","footprint","inequality")])
```


```{r}
# Plotting the maps separately to include the legend
plot(World["HPI"])
plot(World["life_exp"], main = "Life expectancy")
plot(World["footprint"])
plot(World["inequality"])
```


# Part 2
Altitude data for the USA can be obtained using the raster R package:

```{r}
# load libraries
library(raster)
library(sf)

# Get the altitude data for the USA
usa_elev <- raster::getData('alt', country='USA', level=2)

# Extract the data for Oahu
# Hawaii layer is the fourth file
usa_elev4 = usa_elev[[4]]
# Convert to spatial df
spatial_4 <- as(usa_elev4, "SpatialPixelsDataFrame")

# Extract the data for Oahu
# download the shapefile for Oahu
# https://geoportal.hawaii.gov/datasets/cchnl::age-75-to-84-percent/explore?location=21.483374%2C-157.964360%2C11.58
# Download shapefile zip and rename it to Oahu.zip
unzip("Oahu.zip")

# read in the shapefile
oahu_sf <- st_read("Age-Friendly_City_Tract_Percent.shp")
# crop oahu only
oahu_elev <- crop(spatial_4, oahu_sf)
```

## Part 2.1

Making the altitude data for Oahu into a spatial object with geometry type points. Including R code used and output of the data set to show that it is in fact now a spatial object. Selecting a CRS.
```{r}
# Convert the sf object to a points geometry type
oahu_points <- st_as_sf(oahu_elev, geometry = "POINT",
                        crs = 4326)

# View the data to show that it is now a spatial object
str(oahu_points)
oahu_points
```

## Part 2.2

Using a basemap of your choosing, plotting the altitude points onto a map of Oahu. Allow the points to change colors by altitude.

```{r}
# # Split the geometry column of the points into longitude and latitude
oahu_points2 <- oahu_points
coords <- st_coordinates(oahu_points2)
oahu_points2$longitude <- coords[,1]
oahu_points2$latitude <- coords[,2]

# Confirm crs
sf_43 <- sf::st_transform(oahu_sf[,c(1:10)], 4326)
points_43 <- sf::st_transform(oahu_points2, 4326)
# Plot map
ggplot() + 
  geom_sf(data = sf_43) + 
  geom_point(data = points_43, aes(x = longitude, y = latitude,
                                   color = USA4_msk_alt),
             size = 4) + 
  scale_color_gradient(low = "red", high = "blue") + 
  labs(color = "Altitude (m)",
       title = "Altitude points of Oahu")
```

## Part 2.3

Converting the spatial object into a raster, and plot the raster. Including R code and show that the object is now a raster.

```{r}
# create matrix
mat_rast = as.matrix(data.frame(x=oahu_points2$longitude,
                             y =oahu_points2$latitude,
                             z=oahu_points2$USA4_msk_alt))
# convert spatial object to a raster
oahu_rast <- rasterFromXYZ(mat_rast)
# Plot the raster
plot(oahu_rast, main = "Plot of altitude in Oahu")
```

## Part 2.4

Downloading the data for organic matter at https://gis.ctahr.hawaii.edu/SoilAtlas#downloads. What soil orders of organic matters exist at altitudes of > 500 m? Making a map of the organic matter soil orders for altitudes > 500 m.

```{r}
# Download shapefile zip : organic matter
unzip("OrganicMatter.zip")
# unzip shapefile
unzip('tl_2021_15_tract.zip')

# load libraries
library(ggplot2)
library(rgdal)

# Load the shapefiles of Hawaii and soil organic matter
hawaii_shapefile <- readOGR("tl_2021_15_tract.shp")
# Crop only oahu
ha_oahu <- crop(hawaii_shapefile, oahu_sf)
# load organic matter sp
soil_org_shapefile <- readOGR("OrganicMatter_State.shp")

# Define the desired CRS (longitude-latitude)
longlat_crs <- CRS("+proj=longlat +datum=WGS84 +no_defs")
# Transform the shapefile to the desired CRS
organic_sf <- spTransform(soil_org_shapefile, longlat_crs)

# set crs
crs(oahu_elev) <- "+proj=longlat +datum=WGS84 +no_defs"
crs(organic_sf) <-  "+proj=longlat +datum=WGS84 +no_defs"

# Merge based on coordinates and extent
spdf <- over(oahu_elev, organic_sf)

# Add altitude
soilalt <- cbind(spdf, oahu_elev)

# Add categorization of altitude
soilalt$altitude <- ifelse(soilalt$USA4_msk_alt > 500, ">500m","<500m")

# Rename columns
soilalt$longitude <- soilalt$x
soilalt$latitude <- soilalt$y

# Make to simple feature
soil_alt <- st_as_sf(soilalt, coords = c("longitude","latitude"),
                     crs = 4326)

# Plot the filtered shapefile with soil order as the fill color
soil_alt$longitude <- st_coordinates(soil_alt)[,1]
soil_alt$latitude <- st_coordinates(soil_alt)[,2]
```


```{r}
# Plot map
ggplot() + 
  # get from q2.2
  geom_sf(data = sf_43) + 
  geom_point(data = filter(soil_alt, altitude == ">500m"),
                           aes(x = longitude, y = latitude,
                                   color = SoilOrder),
             size = 4) + 
  labs(color = "Soil Order",
       title = "Altitude points of Oahu and Soil Orders Found")

```


# Part 3

## Part 3.1

Using the R packages tmap or mapview, make the plot in Part 1.3 interactive.

```{r}
# load library
library(mapview)
# plot interactive map
mapview(World, zcol = c("HPI", "life_exp", "footprint", "inequality"),
        layer.name = c("HPI", "Life Expectancy", "Footprint", "Inequality"))

```

## Part 3.2

Using the R packages tmap or mapview, making the plot in Part 2.2 interactive.

```{r}
# Plot the map using map view
library(mapview)
# plot
mapview(points_43, zcol = "USA4_msk_alt",
        layer.name = "Altitude points of Oahu",
        colorkey = TRUE, alpha = 0.1, cex = 6,
        legend = TRUE, margin = 0,
        basemap = sf_43)

```



